---
output: html_document
editor_options: 
  chunk_output_type: console
---


# Getting Set Up

We suggest use of the package **remotes** which will avoid re-installing a package if the latest version is already installed. 

```{r install, eval=FALSE}
if (!requireNamespace("remotes")) install.packages("remotes")

pkgs <- c("dplyr", "quadmesh", "raster", "rgl", "scales", "sf", "sfdct", 
"tidyr", "viridis")

remotes::install_cran(pkgs)
```

Obtain some extra data sets that will be used in this session.

```{r get-data,eval=FALSE}
download.file("https://github.com/mdsumner/geo-comp-graphics-oghub/raw/master/data-raw/geo-comp-graphics.Rdata", 
               "geo-comp-graphics.Rdata", mode = "wb")


load("geo-comp-graphics.Rdata")

```


Miscellaneous functions used to build the examples. 


```{r functions-used-for-plots}
colpal <- function(n = 26, drop = 5) sf::sf.colors(n)[-seq_len(drop)]
plot_values <- function(x) {
  raster::plot(x, axes = FALSE, main = "value", box = FALSE, col = colpal()); 
  text(coordinates(x), label = values(x))
  raster::plot(extent(x), add = TRUE)
}
plot_cells <- function(x) {
  raster::plot(x, axes = FALSE, main = "cell", legend = FALSE, box = FALSE, col = colpal()); 
  raster::plot(extent(x), add = TRUE)
  text(coordinates(x), label = sprintf("[%i]", seq_len(ncell(x))), cex = 0.8)
}
plot_edges <- function(x, main = "") {
  sc <- silicate::SC(spex::polygonize(x))
  e <- silicate::sc_edge(sc)
  v <- silicate::sc_vertex(sc)
  x0 <- e %>% dplyr::inner_join(v, c(".vx0" = "vertex_"))
  x1 <- e %>% dplyr::inner_join(v, c(".vx1" = "vertex_"))
  plot(rbind(x0, x1)[c("x_", "y_")], asp = 1, type = "n", 
       axes = FALSE, xlab = "", ylab = "", main = main)
  graphics::segments(x0$x_, x0$y_, x1$x_, x1$y_, lty = 2)
}
```



Other packages used for illustration must be installed from Github, and so require extra tools. These packages aren't necessary for participants. 

```{r install-github, eval=TRUE,echo=FALSE}
##remotes::install_github(c("hypertidy/anglr", "hypertidy/silicate"))

```

