---
output: html_document
editor_options: 
  chunk_output_type: console
---


# The mesh3d format





Rgl is the OpenGL package in R. 

A classic **computer graphics** data model called *mesh3d*, it's not widely used but is very powerful. You can visualize a `mesh3d` model with `shade3d()`, all the aesthetics, material properties, geometry and topology can be attached to the model
itself as data. 

It supports two kinds of primitives **quads** and **triangles**. 


```{r load-data-03}
load("geo-comp-graphics.Rdata")
```

## Quads


Quads are a funny case, usually carried by two triangles (at least implicitly) but they are an important computer graphics element. 


Previously, we created a raster with effectively the following code. 

```{r raw-raster}
m <- matrix(c(seq(0, 0.5, length = 5), 
            seq(0.375, 0, length = 4)), 3)

x <- seq(1, nrow(m)) - 0.5
y <- seq(1, ncol(m)) - 0.5
rast <- raster::raster(list(x = x, y = y, z = m))
```

If we convert that to a quadmesh, the result has vertices (`vb`) and primitives (`ib`). 

```{r quadmesh-03}
library(quadmesh)
qm <- quadmesh(rast)
str(qm)
```


The structure is `vb`, the coordinates of the mesh - these are the
actual corner coordinates from the input raster.  

```{r raster-image}
image(rast, col = colpal())
op <- par(xpd = NA)
text(t(qm$vb), lab = 1:ncol(qm$vb))
par(op)
```


Notice how these are unique coordinates, there's no simple relationship between the cell and its value and its four corners. This is because they are shared between neighbouring cells. The relationship is stored in the `ib` array, this has four rows one for each corner of each cell. There are 9 cells and each has four coordinates from the shared vertex pool. The cells are defined in the order they occur in raster. 

```{r quadmesh-index}
qm$ib
```

It works directly with rgl function, and can be used in more raw form. 

They key point in these plots is that with the exact same 3D geometry, we have the choice of how each primitive is styled. The colour may be a constant every where, or it may vary continously everywhere between vertices. A final choice is that the colour is constant *within a primitive*.  No R 2D plot can easily do the *continuous texture* styling. This also applies to other kinds of textures, later we will apply an image texture to primitives. 


```{r various-quad-aes}

library(rgl)

#clear3d()
shade3d(qm, col = "firebrick")
rglwidget(reuse = FALSE)


```

```{r various-quad-aes2}
clear3d()
quads3d(t(qm$vb)[qm$ib,], col = c("firebrick", "dodgerblue")[qm$ib %% 2 + 1])
rglwidget()
clear3d()
quads3d(t(qm$vb)[qm$ib,], col = rep(c("firebrick", "dodgerblue"), each = 4))
rglwidget()
```

The situation for triangles is much the same, but we have `it` for the triangle index rather than `ib` for the quad index. In both cases the geometry is in the `vb` matrix. Models can have both quads and triangles, using the same set of vertices. 




## Triangles

Triangles aren't so readily created, although it is easy to convert a quadmesh version of a raster into triangles. 

For polygons, we need specialist code. There are two algorithm types: 

* ear-clipping (or ear-cutting)
* near-Delaunay methods

The `mesh_polys` object is create from the `two_polys` simple features layer using ear-clipping. 

This is a very simple algorithm, it chooses one corner of the polygon and clips off a triangle at the next vertex, then proceeds around the boundary - the `decido` package on CRAN uses the Mapbox library `earcut` to do this. 

```{r ear-clipping}
library(sf)
plot(two_polys, reset = FALSE)

polygon(t(mesh_polys$vb[, rbind(mesh_polys$it, NA)]))
```


The `delaunay_polys` object is created from `two_polys` using the in-development `anglr` package, on Github only - anglr uses the `RTriangle` package from CRAN to do this. All edges in the polygons are fed to the algorithm which creates all constrained triangles, then this mesh of triangles is processed to determine which polygon (or hole) each belongs to. With `RTriangle` we can also specify properties like maximum triangle area, or "conform to Delaunay criterion". (This is why we always call polygon triangulation near-Delaunay ...). 

This mesh is both Delaunay-conforming and applying a maximum triangle area, so extra vertices have been added that weren't in the original polygons. 

```{r delaunay-constrained}
plot(two_polys, reset = FALSE)
polygon(t(delaunay_polys$vb[, rbind(delaunay_polys$it, NA)]))
```


## 3D polygons

If we don't have triangles (or quads), we can't plot surfaces or polygons in 3D at all. 

```{r plot-poly3d}
clear3d()
shade3d(mesh_polys)
rglwidget()
```

But, unless we modify the material properties of the surface, or update the geometry in the 3rd dimension we cannot tell the difference between these different mesh polygons. 

```{r plot-del3d}
clear3d()
shade3d(delaunay_polys)
rglwidget()
```


Let's push the Delaunay version up in Z and add some noise. 

```{r delaunay-noise}
clear3d()
delaunay_polys$vb[3,] <- 0
delaunay_polys$vb[3,] <- delaunay_polys$vb[3,] + 
    runif(ncol(delaunay_polys$vb), 1, 1.5)
shade3d(mesh_polys, col = "grey")
wire3d(delaunay_polys)
rglwidget()

```




The primary means to create this format from a raster is for 3D plotting, but because we have access to the coordinate directly it provides other uses. We can transform the coordinates (i.e. a map projection) or manipulate them and augment the Z value (for example) in flexible ways. 

(The usual way of driving rgl grid surfaces is `rgl.surface` but this is limited to the centre-point interpretation only - more than the x, y, z list interface as image() is, i.e. it can take an individual x,y pair for every cell, but it cannot properly represent the cell-as-area as image can. For this we need to use `shade3d`, and actual `meshd3d` types in `rgl`). 


## rgl miscellanea

Most examples around use `rgl.surface`, but I am less familiar with that. The `thing3d()` are the higher-level functions in rgl, and the `rgl.thing()` functions are lower-level (recommended not to mix them in usage). 

**rayshader** in particular, has extremely compelling outputs, but it uses the lower level `rgl.surface` and doesn't maintain the geographic coordinates, so I see it mostly as a texture-generator (but watch its development!). 


**rgl.surface** can take X and Y *matrices*, so you can trivially reproject these data
and wrap them around a sphere - I only learnt this recently. 


**WATCH OUT**

The single most confusing thing I found with `mesh3d` was *homogeneous coordinates*. This is a fourth value on each vertex, making it `X, Y, Z, H` and because it's stored transpose, the `ib` and the `vb` matrices have the same number of rows (4). The `H` loosely corresponds to "zoom", for our purposes set `H = 1`. (If set to 0 no data will be visible, and a package of mine had this bug until last week.)



