
# Transforming spatial data to 3D forms

* What tools exist to convert data to meshes in R? 
* What forms of mesh data are available? 

The key package is rgl, the OpenGL package for R. This tends to be very low level, but has improved helpers and is becoming easier to use. The key functions are `plot3d()`, `lines3d()` and `points3d()` and these are used just like the 2D base funtions in R (with the '3d' on the end). 

For polygons we first need to create triangles, and then use `triangle3d()`. For rasters we can create triangles or quads, and use `quads3d()`. There is also a mesh type object, `mesh3d` to store data with material properties. To plot `mesh3d` we use the `shade3d()` function. 

## Tools

Converting points to 3D is really trivial, we only need the X-Y-Z coordinates, and we can set properties directly on them. 

Converting lines is also relatively easy, but here we might use a *path* model, sequential lists of coordinates that imply a connected line, or a *segments* model where every single individual line segment is specified separately. The segment model is really a mesh, because we have the opportunity to store every coordinate just once and refer to segments by their index. 

Converting polygons is much harder, and there are multiple ways to do it. The main division is by near-Delaunay triangulation, or by ear-cutting algorithms. 

* near-Delaunay - high quality triangulations, suitable for surface modelling
* ear-cutting - low quality triangulation, suitable for flat surfaces

The main tools available for geospatial data: 

* **quadmesh** (on CRAN) - convert any raster to mesh3d quads, optionally with image textures
* **anglr** (not on CRAN) - convert any sp, sf lines or polygons to mesh3d triangles

The `rgl` package has ear-cutting in `triangulate3d()`. 

The `sfdct` package (via `RTriangle`) has near-Delaunay in `ct_triangulate()`

The `decido` package has ear-cutting in `earcut()`. 

Converting raster data is available in various ways. 

- `rgl::surface3d()`, this is analogous to the old base `image()` function that can take `x, y, z = matrix`
- `quadmesh::quadmesh()`, converts raster, matrix, some stars to `mesh3d`
- rayshader determines elevation colouring from a matrix, and has many plotting helpers




## Quadmesh. 


```{r, quadmesh, eval=FALSE}
library(quadmesh)
quadmesh(anyRasterDEM, texture = anyRasterRGB)
rgl::shade3d()

## play with aspect3d, light3d, ...

```

Triangulations, `sfdct` is no good because it's very inefficient. `sf` is simply not suitable for mesh (a.k.a. *indexed*) forms of data. 


### comparison of discrete elevation (extrusion) vs continuous


http://rpubs.com/cyclemumner/geomesh-r



## Image textures

This section requires a Mapbox API key to run from scratch - and can't be shared, so it will simply be illustrated. 


Munster is here. 


```{r ceramic, eval=FALSE, echo=FALSE}
loc <- cbind(7 + 37.5/60, 51 + 57.6/60)
buffer <- 9000
library(ceramic)

munster_elev <- cc_elevation(loc, buffer, zoom = 12)
munster_imag <- cc_location(loc, buffer, zoom = 14)

```

The process is a little complicated, but the key is that the elevation raster provides a space in which to map other data too. In this case the raster is in Mercator, and the image is as well. Under the hood we have to map the spatial part of the image into 0, 1, 0, 1 space of a PNG image, but in the final event we get a scene where real-world data may be added directly. 

```{r ceramic-cache, eval=FALSE, include=TRUE}
munster_qmesh <- quadmesh::quadmesh(munster_elev, texture = munster_imag)

rgl::shade3d(munster_qmesh, lit = FALSE); rgl::aspect3d(1, 1, .05)

## more exact location of THIS building
merc <- sf::sf_project(cbind(7.59551, 51.96922), from = "+init=epsg:4326", to = raster::projection(munster_elev))
rgl::lines3d(cbind(merc[c(1, 1), ], matrix(c(0, 1000))))
```




## Various tools

* `rgl::triangulate` and `decido::earcut` will triangulate polygons with holes but only suitable for plane-filling, because ugly triangles, no control over size and shape. 

* `RTriangle::triangulate` (and `sfdct::ct_triangulate`) do high-quality "near-Delaunay" triangulations

* `quadmesh::quadmesh()` to create rgl-ready mesh3d from a raster

* `mapview::cubeView()` does very compelling interactive raster-cube visualization

* `SymbolixAU/mapdeck()` an alternative 3D viewer using Uber's `deck.gl`

* `tylermorganwall/rayshader`

* `Rvcg` package on CRAN

Side note: there are many triangulation algorithms and many packages in R, but we need *constrained triangulation* to preserve all input edges - only a handful can do that, and RTriangle is the king (with a problematic license). 



* `hypertidy/silicate`, `hypertidy/anglr` - these are evolving together 

* https://github.com/MilesMcBain/gis_vs_web3D 

* `coolbutuseless/threed`


There are a few example data packages:

.obj format: https://github.com/odedstein/meshes/tree/master/objects/koala



## Triangles or quads from geospatial

This is my work-in-progress approach to meshing any data structure. 


```{r, eval=FALSE}
## devtools::install_github("hypertidy/anglr")
## devtools::install_github("hypertidy/silicate")
library(anglr)  
triangles <- copy_down(TRI(anySFpolygon), anyRasterDEM)

mesh <- plot3d(triangles)

```









